---
title: "`HMMoce`: an `R` package for improved analysis of marine animal movement data using hidden Markov models" 
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, 
                      message=FALSE, cache=TRUE,
                      progress = TRUE, verbose = FALSE)

```

**Camrin D Braun, Benjamin Galuardi, Simon R. Thorrold**

## Summary
 1. While the number of marine animals being tagged and tracked continues to grow, current satellite tracking techniques largely constrain meaningful inference to largescale movements of surface-dwelling species and are inherently prone to significant error. The paradigm in fish tracking is to use light and sea surface temperatures to estimate animal movements, largely disregarding the 3-D data collected by the tag.
 2. Oceanographic data and high-resolution models are constantly improving and are becoming widely available and accessible. This data can be leveraged, along with 3-D tag data, to improve estimates of animal movements. We use a likelihood approach for these observations and convolve them with a theoretical movement model in a hidden Markov framework that accomodates state-switching. This discretized approach also provides efficient handling of grid-based oceanographic data and likelihood surfaces generated within the package.
 3. We present an open-source `R` package, `HMMoce`, that uses a novel state-space HMM approach to improve position estimates derived from electronic tags using three-dimensional oceanographic data.
 4. We demonstrate `HMMoce` with example blue shark (*Prionace glauca*) data that is included in the package. Our findings illustrate how our software leverages all available tag data, along with oceanographic information, to improve position estimates of tagged marine species.
 
## Installation Instructions
`HMMoce` can be installed from CRAN from within `R` using `install.packages('HMMoce')`.

## Examples
For an example use of the package, please see the vignette using `vignette('HMMoce')`.
